---
title: "Presentation1"
author: "Pouria"
date: "1/26/2022"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Shiny Presentation

In this lab, we will study regression with 2 predictors, one continuous, and one qualitative, using the
Credit data from the package ISLR

1. Build a linear model of balance as a function of student status. Compare with a t-test. What do you conclude?

2. Now build a model with balance as a function of income. What do you conclude?

3. Next we will build a model with both regressors, student status and income.

a. Plot the data, using different markers for student status

b. Develop a model corresponding to Figure 3.7 left. Plot the model with the data, as in a). What do you conclude?

c. Now develop and plot the full model corresponding to Figure 3.7. Cpmpare the result with those of an ANCOVA analysis. What do you conclude?

## Interactive Plot

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## R Output

```{r cars}
summary(cars)
```


