---
title: "Lab Week 10 - \n PCA"
author: "Pouria"
date: "4/3/2022"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, include=T)
```

```{r, include=F}
library(ggplot2)
library(ISLR)
library(psych)
library(GGally)
library(ggbiplot)
```



## Lab 1 : USArrest Dataset - Pg. 401

## Row Names

```{r}
row.names(USArrests)

```

## Variable Names

```{r}
names(USArrests)
```

## First and second moments
```{r}
apply(USArrests, 2, mean)
sapply(USArrests, mean)
sapply(USArrests, var)
```

## Perform PCA
```{r}
pr.out <- prcomp (USArrests, scale = T)

names(pr.out)
```

## Summary of the output
```{r}
summary(pr.out)
```

## Plot the entire output

```{r}
plot(pr.out)
```



## `$center` and `$scale`

```{r}
pr.out$center

```

```{r}
pr.out$scale
```

## Rotation matrix

```{r}
pr.out$rotation
```

## Biplot

```{r}
biplot(pr.out, scale = 0)
```

## Biplot to match the plot in the book

```{r}
pr.out$rotation=-pr.out$rotation
pr.out$x=-pr.out$x
biplot (pr.out , scale =0)
```


## Use `$sdev` to obtain variance explained by each PC

```{r}
pr.var <- pr.out$sdev^2
pr.var
```


```{r}
pve <- pr.var/sum(pr.var)
pve.cum <- cumsum(pve)
pve.cum
```

## Plot the cumulative variance explained

```{r}

df.var <- data.frame(pve.cum, n.comp = 1:length(pve.cum))

ggplot(df.var) +
  geom_point(aes(x = n.comp, y = pve.cum)) +
  geom_line(aes(x = n.comp, y = pve.cum)) +
  xlab("No. of PCs") +
  ylab("Proportion of Variance Explained")
```





## Lab 3: NCI60 Data - Pg. 407 


## NCI60 Dataset

```{r}
nci.labs <- NCI60$labs
nci.data <- NCI60$data


```


## Investigate the correlations for the first few variables

```{r, message=F, warning=F}
ggpairs(data.frame(nci.data[, 1:5]))
```

## Using `pairs.panels()` from `psych` library

```{r, echo=T}
pairs.panels(nci.data[,1:5],
             gap=0,
             bg = c("red", "yellow", "blue", "green", "black", "purple", "cyan")[factor(nci.labs)],
             pch=21)
```


## Perform PCA

```{r}
pca.out <- prcomp(nci.data, scale. = T)
```



## Plot the pairs or correlation plot for PCs

```{r}
pairs.panels(pca.out$x[,1:5],
             gap=0,
             bg = c("red", "yellow", "blue", "green", "black", "purple", "cyan")[factor(nci.labs)],
             pch=21)


```


## Score Plots (Base R)

```{r}
Cols=function (vec){
  cols=rainbow (length(unique(vec)))
  return(cols[as.numeric (as.factor(vec))])
  }
```

```{r}
par(mfrow=c(1,2))
plot(pca.out$x[,1:2], col=Cols(nci.labs), pch=19,
xlab="Z1",ylab="Z2")
plot(pca.out$x[,c(1,3)], col=Cols(nci.labs), pch=19,
xlab="Z1",ylab="Z3")
```

## Score Plots (ggplot)

```{r}
df.pca.NCI <- data.frame(Z1 = pca.out$x[,1],
                         Z2 = pca.out$x[,2],
                         Z3 = pca.out$x[,3],
                         Cancer.Type = nci.labs)


ggplot(df.pca.NCI) +
  geom_point(aes(Z1, Z2, col = Cancer.Type), size = 2) +
  theme_bw()
  

```

```{r}
ggplot(df.pca.NCI) +
  geom_point(aes(Z1, Z3, col = Cancer.Type), size = 2) +
  theme_bw()
```


## Summary of the output

```{r}
summary(pca.out)
```



## Plot Variance Explained Using `summary()`

```{r}
ggplot() +
  geom_point(aes(1:length(summary(pca.out)$importance[3,]) ,
                 summary(pca.out)$importance[3,])) +
  xlab("Number of PCs") +
  ylab("Cumulative PVE")
```



## Biplot in ggplot 

Use `ggbiplot()` as an alternative to base R `biplot()`

```{r, include=T, echo=T}
# install.packages("remotes")
# remotes::install_github("vqv/ggbiplot")
# library(ggbiplot)
ggbiplot(pca.out,
         obs.scale = 1,
         var.scale = 1,
         groups = nci.labs,
         ellipse = TRUE,
         circle = TRUE,
         ellipse.prob = 0.68) +
  scale_color_discrete(name = '') +
  theme(legend.direction = 'horizontal',
               legend.position = 'top')
```




