---
title: 'Lab Week 03: Linear Regression - Part 2'
author: "BKN"
date: "1/31/2022"
output:
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Lab Prompt

In this lab, we will study regression with 2 predictors, one continuous, and one qualitative, using the Credit data from the package `ISLR`

1. Build a linear model of `balance` as a function of `student` status. Compare with a t-test. What do you conclude?
2. Now build a model with `balance` as a function of `income`. What do you conclude?
3. Next we will build a model with both regressors, `student` status and `income`.
    
    3.a) Plot the data, using different markers for `student status`
    
    3.b) Develop a model corresponding to Figure 3.7 left. Plot the model with the data, as in a). What do you conclude?
    
    3.c) Now develop and plot the full model corresponding to Figure 3.7. Compare the result with those of an ANCOVA analysis. What do you conclude?



```{r , echo=FALSE}
library(ISLR)
library(ggplot2)
attach(Credit)

```

## Lab 1

Model `Balance` as a function of `Student`

```{r 1}
# Model balance as a function of Student
```


## Lab 1
Compare with t-test

```{r 1 t-test}
# Compare the results with those from a t-test
```


## Lab 2

Model `Balance` as a function of `Income`

```{r 2}
# Model Balance as a function of Income
```


## Lab 3(a)

Plot the data points

```{r 3a}

```

## Lab 3(b)

Model `Balance` as a function of both `Student` and `Income`

```{r 3b}
# Model Balance as a function of Student and Income
```


## Lab 3(b)

Or just simply use `Student` with no need for introducing a dummy variable.

```{r 3b how does R imply categorical variables?}
# or just use Student; R automatically does hot-encoding

```


## Lab 3(b)

Plot `Balance` as a function of `Student` and `Income` (Similar to Fig. 3.7(Left))

```{r 3b plot}

```


## Lab 3(c)

Model `Balance` as a function of both `Student`, `Income` and their interaction

```{r 3c model}
# Model Balance as a function of Student, Income and their interaction
```


## Lab 3(c)

Compare with ANCOVA

```{r 3c ANCOVA}
# Compare the results with those from ANCOVA
```



## Lab 3(c)

Plot the full model (Similar to 3.7(Right))

```{r Plot 3c}

```

