---
title: "General Regression: Ridge, Lasso, PCR & PLS"
author: "Pouria"
date: "4/18/2022"
output: slidy_presentation

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

## Prompt

Looking at the `College` data set, we intend to predict the number of applications received, `Apps`, using the other variables.

a) Split the data set into a training set and a test set with 20% of the data held out as the test set. 

b. Fit a PCR model on the training set, with M chosen by cross-validation. Report the test error obtained, along with the value of M selected by cross-validation.

c) Fit a PLS model on the training set, with M chosen by cross-validation. Report the test error obtained, along with the value
of M selected by cross-validation.

d) Fit a linear model using least squares on the training set, and report the test error obtained.

e) Fit a ridge regression model on the training set, with $\lambda$ chosen
by cross-validation. Report the test error obtained.

f) Fit a lasso model on the training set, with λ chosen by cross-validation. Report the test error obtained, along with the number of non-zero coefficient estimates.

g) Comment on the results obtained. How accurately can we predict the number of college applications received? Calculate test R-squared values. Is there much
difference among the test errors resulting from these five approaches?


## Libraries

```{r, warning=F, message=F}
library(ISLR)
library(glmnet)
library(pls)
library(ggplot2)
library(GGally)
library(dplyr)
library(tidyr)
```


## Set seed

```{r}
set.seed(11)
```


## (a) Split the data

Split the data set into a training set and a test set. Hold 20% of the data out for test. 

```{r}
sum(is.na(College))

College[, "Apps"] = College[, "Apps"] + rnorm(dim(College)[1], 1000, 5000)

# College[, "Apps"] = College[, "Apps"] + rnorm(dim(College)[1], 1000, 10000)


# College1 <- data.frame(College,
#                        var1 = rnorm(dim(College)[1], 1000,5000),
#                        var2 = rnorm(dim(College)[1], 2000, 1000))

train.size = round(dim(College)[1] *.8)
train = sample(dim(College)[1], train.size)
test = -train

College.train <- College[train, ]
College.test  <- College[test, ]

dim(College.train)
train.size
var(College[, "Apps"])
```





## (b) Principal Component Regression (PCR)

- Fit a PCR model on the training set, with number of PCs chosen by cross-validation. Visualize the results.

- Report the test error corresponding with the number of PCs selected from above.

- Report the variance explained in both predictors and response spaces across different number of PCs.


## (b) PCR

1. Perform PCR on the training set and visualize MSE across number of PCs using cross-validation.


```{r}
pcr.cv = pcr(Apps~., data=College.train, scale=T, validation="CV")

validationplot(pcr.cv, val.type="MSEP")

```

## (b) PCR

- Can you manually plot what `validationplot()` gave you above?

*Hint*: Look into `$validation$pred` and use index `[,1,]` to access the predictions per number of components. And remember you are working with a data frame so `apply()` can come in handy for any type of computation across different variables.


```{r}
res.pcr <- apply(pcr.cv$validation$pred[,1,], 2, function(x) mean((x-College.train[,"Apps"])^2))

dim(pcr.cv$validation$pred)

ggplot() +
  geom_line(aes(1:pcr.cv$ncomp, res.pcr)) +
  geom_point(aes(1:pcr.cv$ncomp, res.pcr)) +
  xlab("Number of PCs") +
  ylab("MSEP") +
  scale_x_continuous(breaks = seq(1,pcr.cv$ncomp, 1))
```

## (b) PCR

Use `selectNcomp()` to select the number of PCs that correspond to 1 SE.

```{r}
N.PC <- selectNcomp(pcr.cv, method = "onesigma", plot = TRUE)

```



## (b) PCR

2. Report the test MSE for PCR with the number of PCs selected above. 

```{r}
pcr.pred = predict(pcr.cv, College.test, ncomp=N.PC)

pcr.mse <- mean((College.test[, "Apps"] - pcr.pred)^2)

pcr.mse
```


## (b) PCR

3. Report a summary of the PCR model and visualize the variance explained in predictors space for different number of PCs.

- Summary

```{r}
summary(pcr.cv)
```

## (b) PCR

3. Report a summary of the PCR model and visualize the variance explained in both predictors and response variable spaces for different number of PCs.

- Visualize the variance explained

*Hint 1*: You may access the variance in predictors through `$Xvar`. Remember to use cumulative sum.

*Hint 2*: Also, `$residuals[,1,]` will give you the residuals for each number of PCs. Note that residuals and variance in response variable `Apps` are related. Variance explained in response variable space is the same as $R^2$.


```{r}
XVar <- cumsum(pcr.cv$Xvar) / pcr.cv$Xtotvar

Yres <- apply(pcr.cv$residuals[,1,], 2, function(x) 1 - mean(x^2)/mean((College.train[,"Apps"]-mean(College.train[,"Apps"]))^2))

pcr.explained.df <- data.frame(N.Comp = 1:pcr.cv$ncomp, X = XVar, Y = Yres) %>%
  pivot_longer(cols = 2:3, names_to = "Variance", values_to = "Var.Values")


ggplot(pcr.explained.df) +
  geom_line(aes(N.Comp,Var.Values, col = Variance), size = 1.1) +
  geom_point(aes(N.Comp,Var.Values, col = Variance)) +
  scale_color_manual(values = c("red", "black")) +
  xlab("Number of Principal Components") +
  ylab("Variance Explained") +
  scale_x_continuous(breaks = seq(1,pcr.cv$ncomp, 1)) +
  scale_y_continuous(breaks = seq(0,1, .1))
```





## (c) Partial Least Square Regression (PLSR)

- Fit a PLS model on the training set, with number of components chosen by cross-validation. 

- Report the test error obtained, along with the value
of M selected by cross-validation.

- Report the variance explained in both predictors and response spaces across different number of PCs.

## (c) PLSR

1. Perform PLS on the training set and visualize MSE across number of components using cross-validation.

```{r}
pls.cv = plsr(Apps~., data=College.train, scale=T, validation="CV")
validationplot(pls.cv, val.type="MSEP")
```

## (c) PLSR

- Plot manually

```{r}
res.pls <- apply(pls.cv$validation$pred[,1,], 2, function(x) mean((x-College.train[,"Apps"])^2))

ggplot() +
  geom_line(aes(1:pls.cv$ncomp, res.pls)) +
  geom_point(aes(1:pls.cv$ncomp, res.pls)) +
  xlab("Number of PCs") +
  ylab("MSEP") +
  scale_x_continuous(breaks = seq(1,pls.cv$ncomp, 1))

```

## (c) PLSR

Use `selectNcomp()` to select the number of PCs that correspond to 1 SE.

```{r}
N.PC <- selectNcomp(pls.cv, method = "onesigma", plot = TRUE)

```


## (c) PLSR

2. Report the test MSE for PLS with the number of PCs selected above. 

```{r}
pls.pred = predict(pls.cv, College.test, ncomp=N.PC)
pls.mse <- mean((College.test[, "Apps"] - pls.pred)^2)
pls.mse
```


## (c) PLSR
3. Report a summary of the PCR model and visualize the variance explained in predictors space for different number of PCs.

- Summarize

```{r}
summary(pls.cv)
```

## (c) PLSR

3. Report a summary of the PCR model and visualize the variance explained in both predictors and response variable spaces for different number of PCs.

- Visualize the variance explained

```{r}
XVar <- cumsum(pls.cv$Xvar) / pls.cv$Xtotvar

Yres <- apply(pls.cv$residuals[,1,], 2, function(x) 1 - mean(x^2)/mean((College.train[,"Apps"]-mean(College.train[,"Apps"]))^2))

pls.explained.df <- data.frame(N.Comp = 1:pls.cv$ncomp, X = XVar, Y = Yres) %>%
  pivot_longer(cols = 2:3, names_to = "Variance", values_to = "Var.Values")


ggplot(pls.explained.df) +
  geom_line(aes(N.Comp,Var.Values, col = Variance), size = 1.1) +
  geom_point(aes(N.Comp,Var.Values, col = Variance)) +
  scale_color_manual(values = c("red", "black")) +
  xlab("Number of Principal Components") +
  ylab("Variance Explained") +
  scale_x_continuous(breaks = seq(1,pls.cv$ncomp, 1)) +
  scale_y_continuous(breaks = seq(0,1, .1))
```

## (b) & (c)

Compare the results from PCR and PLS. How are the values and plots in (c) different from those obtained for PCR in (b). How can this difference be rationalized?





## (d) Fit a linear model

Fit a linear model using least squares on the training set, and report the test error obtained.

```{r}
lm.fit = lm(Apps~., data=College.train)
summary(lm.fit)

```

```{r}
lm.pred = predict(lm.fit, College.test)
ols.mse <- mean((College.test[, "Apps"] - lm.pred)^2)

ols.mse
```

## (e) Ridge Regression

- Fit a ridge regression model on the training set, with $\lambda$ chosen
by cross-validation. 

- Report the test error obtained.

## (e) Ridge Regression

1. First, turn the data into matrix to take care of categorical variables; One method to do this is `model.matrix` for both test and train sets. Note that this should be used for `cv.glmnet()`.

```{r}
train.mat = model.matrix(Apps~., data=College.train)
test.mat = model.matrix(Apps~., data=College.test)

```

## (e) Ridge Regression

2. Now, Find the best $\lambda$.

```{r}
cv.ridge = cv.glmnet(train.mat, College.train[, "Apps"], alpha=0)
lambda.best = cv.ridge$lambda.min
lambda.best

```


## (e) Ridge Regression

3. Now, report the test error using Ridge with best $\lambda$.

```{r}
ridge.pred <- predict(cv.ridge, newx=test.mat, s=lambda.best)

ridge.mse <- mean((College.test[, "Apps"] - ridge.pred)^2)

ridge.mse

```

## (f) Lasso Regression

- Fit a lasso model on the training set, with $\lambda$ chosen by cross-validation. 
- Report the test error obtained, ...

  - along with the coefficients.


## (f) Lasso Regression

1. Now, Find the best $\lambda$.

```{r}
cv.lasso = cv.glmnet(train.mat, College.train[, "Apps"], alpha=1)
lambda.best = cv.lasso$lambda.1se
lambda.best

```


## (f) Lasso Regression

2. Now, report the test error using Lasso with best $\lambda$.

```{r}

lasso.pred <- predict(cv.lasso, newx=test.mat, s=lambda.best)

lasso.mse <- mean((College.test[, "Apps"] - lasso.pred)^2)

lasso.mse

```

## (f) Lasso Regression

3. Report the coefficients of the best Lasso model with best $\lambda$.

```{r}
# lasso.out <- glmnet(model.matrix(Apps~., data=College), College[, "Apps"], alpha=1)
predict(cv.lasso, type="coefficients", s=lambda.best)

```


## (g) Compare Models

Comment on the results obtained. 

- How accurately can we predict the number of college applications received? 

- Is there much difference among the test errors resulting from these five approaches? Visualize using bar-plots.

- Calculate test R-squared values and compare. Visualize using bar-plots.



## (g) Compare Models

1. Is there much difference among the test errors resulting from these five approaches? Visualize using bar-plots.

```{r}
df.mse <- data.frame(Model = c("LM", "Ridge", "Lasso", "PCR", "PLS"),
                     MSE = c(ols.mse, ridge.mse, lasso.mse, pcr.mse, pls.mse))

ggplot(df.mse) +
  geom_col(aes(Model, MSE), width = .3) +
  ylab("Test MSE") +
  theme_bw() 
```

## (g) Compare Models

2. Calculate test R-squared values and compare. Visualize using bar-plots.


```{r}
test.avg = mean(College.test[, "Apps"])

TSS = mean((College.test[, "Apps"] - test.avg)^2)

R2.calc <- function(RSS, TSS){
  return (1 - RSS/TSS)
}

ols.r2 <- R2.calc(ols.mse, TSS)
ridge.r2 <- R2.calc(ridge.mse, TSS)
lasso.r2 <- R2.calc(lasso.mse, TSS)
pcr.r2 <- R2.calc(pcr.mse, TSS)
pls.r2 <- R2.calc(pls.mse, TSS)

df.r2 <- data.frame(Model = c("LM", "Ridge", "Lasso", "PCR", "PLS"),
                     r2 = c(ols.r2, ridge.r2, lasso.r2, pcr.r2, pls.r2))

ggplot(df.r2) +
  geom_col(aes(Model, r2), width = .3) +
  ylab("Test R2") +
  scale_y_continuous(breaks = seq(0,1, .1), expand = c(0,1)) +
  theme_bw()

```



